ggalluvial: Layered Grammar for Alluvial Plots

### 2019-12-05 paper.rmd
try to rm figure caption

---
title: "ggalluvial: Layered Grammar for Alluvial Plots"
tags:
  - R
  - ggplot2
  - alluvial plots
  - statistical graphics
  - data visualization
  - repeated measures data
  - categorical data
authors:
  - name: Jason Cory Brunson
    orcid: 0000-0003-3126-9494
    affiliation: "1"
affiliations:
 - name: Center for Quantitative Medicine, UConn Health
   index: 1
date: 2019-12-05
bibliography: paper.bib
output: md_document
---

@@

```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6}
library(ggalluvial)
combos <- data.frame(
  stat = c(
    rep("stratum", 2),
    rep("alluvium", 4),
    "flow",
    # text geom
    c("stratum", "alluvium", "flow"),
    # errorbar and pointrange geoms
    "stratum", "alluvium", "flow"
  ),
  geom = c(
    c("stratum", "lode"),
    c("stratum", "alluvium", "lode", "flow"),
    "flow",
    rep("text", 3),
    "errorbar", "pointrange", "pointrange"
  ),
  pairing = c(
    c("default", "alluvial"),
    c("alluvial", "default", "default", "alluvial"),
    "default",
    rep("external", 6)
  )
)
levs <- c("stratum", "alluvium", "lode", "flow",
          "text", "errorbar", "pointrange")
combos <- transform(
  combos,
  stat = factor(stat, levs),
  geom = factor(geom, levs),
  pairing = factor(pairing, c("default", "alluvial", "external"))
)
ggplot(combos, aes(axis1 = stat, axis2 = geom)) +
  scale_x_discrete(limits = c("stat", "geom"), expand = c(.1, .05)) +
  geom_alluvium(aes(fill = pairing)) +
  scale_fill_brewer(type = "qual") +
  geom_stratum() +
  geom_text(stat = "stratum", infer.label = TRUE) +
  theme_minimal()
```

@@

**Incomplete longitudinal data:**
Alluvial plots very clearly delineate times at which longitudinal data are censored, discontinued, or otherwise missing: Certain strata, or the alluvia or flows connecting them, are present at one time point but absent at a previous or future one.
@Seekatz2018 use this feature to include in one alluvial plot a sample of _Clostridium difficile_--infected (CDI) patients who had their infections ribotyped at multiple times. Patients were classified by dominant ribotype, and the alluvial plot showcased variability in this dominant type. While all 32 patients had at least two samples taken, only 3 had four, so that the bar plots shortened with each time point.
@Sjoding2019 use a similar plot to trace patient groups receiving mechanical ventilation based on discretized tidal volumes, including a grey stratum for patients discontinued from intubation rather than omitting them entirely from the plot.
<!--
Even in cases of intermediate missingness, when a case is lost, or its value diminished to zero, between two time points at which it is positively measured, alluvial diagrams can preserve the continuity of such cases by shrinking their ribbon widths to zero at the intermediate point. (See the example in the `stat_alluvium()` examples using the babynames package.)
-->

>>>>

---
title: "ggalluvial: Layered Grammar for Alluvial Plots"
tags:
  - R
  - ggplot2
  - alluvial plots
  - statistical graphics
  - data visualization
  - repeated measures data
  - categorical data
authors:
  - name: Jason Cory Brunson
    orcid: 0000-0003-3126-9494
    affiliation: "1"
affiliations:
 - name: Center for Quantitative Medicine, UConn Health
   index: 1
date: 2019-12-05
bibliography: paper.bib
---

@@

```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.cap=NA}
library(ggalluvial)
combos <- data.frame(
  stat = c(
    rep("stratum", 2),
    rep("alluvium", 4),
    "flow",
    # text geom
    c("stratum", "alluvium", "flow"),
    # errorbar and pointrange geoms
    "stratum", "alluvium", "flow"
  ),
  geom = c(
    c("stratum", "lode"),
    c("stratum", "alluvium", "lode", "flow"),
    "flow",
    rep("text", 3),
    "errorbar", "pointrange", "pointrange"
  ),
  pairing = c(
    c("default", "alluvial"),
    c("alluvial", "default", "default", "alluvial"),
    "default",
    rep("external", 6)
  )
)
levs <- c("stratum", "alluvium", "lode", "flow",
          "text", "errorbar", "pointrange")
combos <- transform(
  combos,
  stat = factor(stat, levs),
  geom = factor(geom, levs),
  pairing = factor(pairing, c("default", "alluvial", "external"))
)
ggplot(combos, aes(axis1 = stat, axis2 = geom)) +
  scale_x_discrete(limits = c("stat", "geom"), expand = c(.1, .05)) +
  geom_alluvium(aes(fill = pairing)) +
  scale_fill_brewer(type = "qual") +
  geom_stratum() +
  geom_text(stat = "stratum", infer.label = TRUE) +
  theme_minimal()
```

@@

**Incomplete longitudinal data:**
Alluvial plots very clearly delineate times at which longitudinal data are censored, discontinued, or otherwise missing: Certain strata, or the alluvia or flows connecting them, are present at one time point but absent at a previous or future one.
@Seekatz2018 use this feature to include in one alluvial plot a sample of _Clostridium difficile_--infected (CDI) patients who had their infections ribotyped at multiple times. Patients were classified by dominant ribotype, and the alluvial plot showcased variability in this dominant type. While all 32 patients had at least two samples taken, only 3 had four, so that the bar plots shortened with each time point.
@Sjoding2019 use a similar plot to trace patient groups receiving mechanical ventilation based on discretized tidal volumes, including a grey stratum for patients discontinued from intubation rather than omitting them entirely from the plot.
